---
title: "PCA_for_STRADLCog"
author: "Holly Spence"
date: "26/05/2021"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r}
library("factoextra")
library("tidyverse")
library("gridExtra")
library("Hmisc")
library("corrplot")
```

Tidyverse package allow referral to all variables instead of using column names. 

Create data frame just for cognition data
```{r}
pca_data<-data %>% select(c(ID, ImmediateRecall:MatrixReasoning_time))
```

Create correlation matrix of cog variables:
```{r}
corrnums<-pca_data %>% filter(complete.cases(.)) %>% select(-ID) %>% cor() #outputs correlation coefficients for those participants with no missing data
corr_ps<-pca_data %>% filter(complete.cases(.)) %>% select(-ID) %>% as.matrix() %>% rcorr(type = "pearson") # outputs correlation p values using pearson correlations
corr_ps
```

Represent this in a correlation plot:
```{r}
M<-corr_ps$r # uses correlation coefficient table to create new object M
p_mat<- corr_ps$p # uses p values from rcorr command
corrplot(M, type = "upper", method = "circle", order = "hclust", p.mat = p_mat, sig.level = 0.05, insig = "blank")
```

*Run the PCA*
First create a table without missing values:
```{r}
pca_data<-pca_data %>% filter(complete.cases(.))
pca_data<-tibble::rowid_to_column(pca_data, "row_num") #need this to create a data frame (tibble function does this) with row numbers so can then merge back to original data frame.
```

Use prcomp function to carry out initial PCA 
```{r}
pcacog<-prcomp(pca_data[,c(3:9)], center = TRUE, scale. = TRUE)
#center argument centers the values around 0 and scale argument scales the variables to have unit variance before the analysis is started
summary(pcacog)
# Flip the first principal component loading
pcacog$rotation[, 1] <- -pcacog$rotation[, 1]

# If you also want to flip the scores accordingly
pcacog$x[, 1] <- -pcacog$x[, 1]
```
Principal components 1-8 each explain a percentage of the variation in the dataset. e.g. PC1 explains 37.86% of total variance, PC2 explains 19.58% variation etc. 

Assess this using the scree plot:
```{r}
fviz_eig(pcacog, addlabels = TRUE)#creates scree plot to look at eigen values
```

Summarise PCA using eigenvalues:
```{r}
loading_matrix<-loadings(pcacog)
fviz_pca_var(pcacog, col.var = "contrib", #shows contribution to PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, xlab = "Principal Component 1 (37.7%)", ylab = "Principal Component 2 (20.5%)")
```

Extract the principal components for each participant to give them a g (general cognition) value. 
Essentially using principal component model to predict a g value for all participants using PC1 which explains the most variation of all the principal components.
```{r}
extractpca<-predict(pcacog, newdata = pca_data)
head(extractpca)
```

Merge these PC scores into the data frame
```{r}
#create data frame of PC matrix
extractpcaasdf<-as.data.frame(extractpca)
extractpcaasdf<-rowid_to_column(extractpcaasdf , "row_num")
#merge cog data and PC df using row num and STRADL ID as identifiers then remove PCs after PC1 
masterpca<-left_join(pca_data, extractpcaasdf, by="row_num")
master_data <- masterpca %>% select(-c(row_num, PC2:PC7))
master_data <- master_data %>% rename(g=PC1)
```

Check that g correlates with individual factors more than they correlate with eachother - e.g.:
```{r}
corr_ps2 <-  master_data  %>% select(-ID) %>% as.matrix() %>%  rcorr(type = 'pearson')
corr_ps2
master_data %>% ggplot(aes(x=g,y=mema))+geom_point(alpha=.4)+theme_bw()
master_data %>% ggplot(aes(x=g,y=digsym),)+geom_point(alpha=.4)+theme_bw()
master_data %>% ggplot(aes(x=g,y=mrtottime),alpha=.4)+geom_point(alpha=.4)+theme_bw()
master_data %>% ggplot(aes(x=mrtotc,y=digsym),alpha=.4)+geom_point(alpha=.4)+theme_bw()

```
